Computationally Efficient Simulation of Queues: The \proglang{R} Package \pkg{queuecomputer}

### 2020-07-17 paper.tex
Rename to match names from Bettina

\RequirePackage{xcolor}
\documentclass[article]{jss}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{amsmath,amssymb,array}
% \usepackage{booktabs}

%%%%%%%%%%%Algorithm Things%%%%%------
% \usepackage{algorithmic}

\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{algorithm}
% \usepackage[section]{placeins}

%%%%%%%%%%%%So I can read in tables%%%%%------
\usepackage{pgfplotstable,filecontents}
\pgfplotsset{compat=1.9}% supress warning

%%%%%%% For the notation table %%%%%%%%%-------
\usepackage{multirow}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Anthony  Ebert\\Queensland University of Technology\\ACEMS \And 
        Paul Wu\\Queensland University of Technology\\ACEMS \AND
        Kerrie Mengersen\\Queensland University of Technology\\ACEMS \And
        Fabrizio Ruggeri\\CNR-IMATI\\Queensland University of Technology\\ACEMS}
\title{Computationally Efficient Simulation of Queues: The \proglang{R} Package \pkg{queuecomputer}}

\Shorttitle{queuecomputer: Computationally Efficient Simulation of Queues}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Anthony Ebert, Paul Wu, Kerrie Mengersen, Fabrizio Ruggeri} %% comma-separated
\Plaintitle{Computationally Efficient Simulation of Queues: The R Package queuecomputer} %% without formatting
% \Shorttitle{\pkg{queuecomputer}: A Capitalized Title} %% a short title (if necessary)

%% an abstract and keywords
\Abstract{
Large networks of queueing systems model important real-world systems such as MapReduce clusters, web-servers, hospitals, call centers and airport passenger terminals. To model such systems accurately, we must infer queueing parameters from data. Unfortunately, for many queueing networks there is no clear way to proceed with parameter inference from data. Approximate Bayesian computation could offer a straightforward way to infer parameters for such networks if we could simulate data quickly enough.

We present a computationally efficient method for simulating from a very general set of queueing networks with the \proglang{R} package \textbf{queuecomputer}. Remarkable speedups of more than 2 orders of magnitude are observed relative to the popular DES packages \pkg{simmer} and \pkg{simpy}. We replicate output from these packages to validate the package.

The package is modular and integrates well with the popular \proglang{R} package \pkg{dplyr}. Complex queueing networks with tandem, parallel and fork/join topologies can easily be built with these two packages together. We show how to use this package with two examples: a call center and an airport terminal. 
}
\Keywords{queues, queueing theory, discrete event simulation, 
operations research, approximate Bayesian computation, \proglang{R}}
\Plainkeywords{queues, queueing theory, discrete event simulation, 
operations research, approximate Bayesian computation, R} %% without formatting
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Anthony Ebert\\
  School of Mathematical Sciences\\
  Science and Engineering Faculty\\
  Queensland University of Technology\\
  Brisbane Queensland 4000, Australia\\
  E-mail: \email{ac.ebert@qut.edu.au}\\
  URL: \url{https://acems.org.au/our-people/anthony-ebert}
}

@@

Performance measures for a given queueing system can often only be derived through simulation. Queues are usually simulated with discrete event simulation (DES) \citep[pg. 226]{insua2012bayesian}. In DES changes in state are discontinuous. The state is acted upon by a countable list of events at certain times which cause the discontinuities. If the occurrence of an event is independent of everything except simulation time it is determined; otherwise, it is contingent \citep{nance1981time}.

Popular DES software packages are available in many programming languages including: the \proglang{R} package \pkg{simmer} \citep{Rpkg_simmer}, the \proglang{Python} \citep{van2011python} package \pkg{simpy} \citep{Ppkg_simpy} and the \proglang{Java}  \citep{gosling2000java} package \pkg{JMT} \citep{Jpkg_JMT}. DES packages are often so expressive that they can be considered languages in their own right, indeed the programming language \proglang{Simula} \citep{dahl1966simula} is a literal example of this.

\pkg{queuecomputer} \citep{Rpkg_queuecomputer} implements an algorithm that can easily be applied to a wide range of queueing systems and networks of queueing systems. It is vastly more computationally efficient than existing approaches to DES. We term this new computationally efficient algorithm queue departure computation (QDC). Computational efficiency is important because if we can simulate from queues quickly, then we can embed a queue simulation within an approximate Bayesian computation (ABC) algorithm \citep{sunnaker_approximate_2013} and estimate queue parameters for very complicated queueing models in a straightforward manner.

In Section \ref{sec:queueing} we review the literature on queueing theory and develop notation used throughout this paper. In Section \ref{sec:QDC} we present the QDC algorithm and compare it to DES. We demonstrate usage of the package in Section \ref{sec:usage}. Details of implementation and usage are discussed in Section \ref{sec:Implementation}. The package is validated in Section \ref{sec:Validation} by replicating results from DES packages \pkg{simpy} and \pkg{simmer}. We compare computed performance measures from the output of a \pkg{queuecomputer} simulation to theoretical results for $M/M/2$ queueing systems. We benchmark the package in Section \ref{sec:Benchmark} and compare computation time with \pkg{simpy} and \pkg{simmer}. Examples in Section \ref{sec:Examples} are used to demonstrate how the package can be used to simulate a call center and an international airport terminal.

@@

Queueing theory is the study of queueing systems and originated from the work of Agner Krarup Erlang in 1909 to plan infrastructure requirements for the Danish telephone system \citep[pg 2]{thomopoulos2012fundamentals}.

A queueing system is defined as follows. Each customer $i = 1,2,\cdots$ has an arrival time $a_i$ (or equivalently an inter-arrival time $\delta_i = a_i - a_{i-1}$, $a_0 = 0$) and an amount of time they require with a server, called the service time $s_i$. Typically a server can serve only one customer at a time. A server which is currently serving another customer is said to be unavailable, a server without a customer is available. If all servers are unavailable when a customer arrives then customers must wait in the queue until a server is available. Detailed introductions to queueing systems can be found in standard texts such as \citet{bhat2015introduction}.

The characteristics of a queueing system are expressed with the notation of \citet{kendall1953stochastic}. This notation has since been extended to six characteristics:
\begin{itemize}
\item $f_{\mathbf{\delta}}$, inter-arrival distribution;
\item $f_{\mathbf{s}}$, service distribution; 
\item $K$, number of servers $\in \mathbb{N}$;
\item $C$, capacity of system $\in \mathbb{N}$; 
\item $n$, customer population $\in \mathbb{N}$; and
\item $R$, service discipline
\end{itemize}

Choices for inter-arrival and service distributions are denoted by ``M" for exponential and independently distributed, ``GI" for general and independently distributed and ``G" for general without the independence assumption. The capacity of the system $C$ refers to the maximum number of customers within the system at any one time\footnote{If the system is at full capacity and new customers arrive, new customers leave the system immediately without being served.}. Customers are within the system if they are being served or waiting in the queue. The customer population $n$ is the total number of customers including those outside of the system (yet to arrive or already departed). The service discipline $R$ defines how customers in the queue are allocated to available servers. The most common service discipline is first come first serve (FCFS). To specify a queueing system, these characteristics are placed in the order given above and separated by a forward slash ``/".

The simplest queueing system is exponential in distribution for both the inter-arrival $\delta_i \overset{iid}{\sim} \mathrm{exp}(\lambda) ~ \forall i \in 1:n$ and service processes $s_i \overset{iid}{\sim} \mathrm{exp}(\mu) ~ \forall i \in 1:n$, where $\lambda$ and $\mu$ are exponential rate parameters. Additionally, $K$ is set to 1, $C$ and $n$ are infinite, and $R$ is FCFS. It is denoted by $M/M/1/ \infty / \infty / FCFS$, which is shortened to $M/M/1$.

Parameter inference for this system was considered first by \citet{clarke1957maximum}, estimators were derived from the likelihood function. This likelihood is later used by \citet{muddapur1972bayesian} to derive the joint posterior distribution. Bayesian inference for queueing systems is summarised in detail by \citet{insua2012bayesian}.

@@

In the case of a $M/M/K$, system equilibrium distributions for performance measures can be derived analytically. These derivations are found in standard queueing theory textbooks \citep{lipsky2008queueing, thomopoulos2012fundamentals}. For instance, the limit probability of $N$ customers in the system $\Prob(N)$  is

\begin{align}
\Prob(0) &= \left[\frac{ (K \rho)^K}{K! (1 - \rho)} + 1 + \sum_{i=1}^{K-1} \frac{(K \rho)^i}{i!}   \right]^{-1} \nonumber \\
\Prob(N) &= \begin{cases} 
\Prob(0) \frac{(K \rho)^n }{N!} & \quad  N \leq K\\
\Prob(0) \frac{(K \rho)^n }{K! K^{N-K}} & \text{otherwise} 
\end{cases} \label{eq:pn}
\end{align}

where $\rho$, the resource utilization, is defined as $\frac{\lambda}{K \mu}$. For an $M/M/K$ system this is equal to the expected number of busy servers divided by the total number of servers $\frac{\E(B)}{K}$ \citep[pg. 451]{cassandras2009introduction}. The expected number of customers in the system is \citep{bhat2015introduction}

\begin{align}
\E(N) = K\rho + \frac{\rho (K\rho)^K P(0)}{K!(1-\rho)^2} \label{eq:en},
\end{align}

and the expected waiting time is

\begin{align}
\E(\mathbf{w}) = \frac{(K \rho)^K P(0)}{K! K \mu(1-\rho)^2} \label{eq:ew}.
\end{align}

If the parameters of $f_\delta$ and $f_{\mathbf{s}}$ are uncertain, then we must turn to predictive distributions for estimates of performance measures, which are computed analytically for M/M/K queues (Equations \ref{eq:en} and \ref{eq:ew}). Predictive distributions of performance measures using Bayesian posterior distributions are derived by \citet{armero1994bayesian, armero1999dealing}.

@@

In a tandem queueing network, customers traverse an ordered series of queues before departing the system. Real examples of such systems include airport terminals, internet services and manufacturing systems. In a parallel network, customers are partitioned into different queueing systems, with their own two vectors $\mathbf{a}$ and $\mathbf{s}$. In a fork/join network a task (another term for customer) is split into a number of subtasks which are to be completed by distinct parallel servers. The task is only complete once all its associated subtasks arrive at the join point. One example of this situation is that of arriving passengers and their bags at an airport.

The difference from the parallel network is that the task can only depart the system once all its associated subtasks have arrived at the join point.

@@

In general, analytic solutions do not exist for dynamic queueing systems \citep{malone1995dynamic, worthington2009reflections}. \citet{green1991some} showed that using stationary queueing systems to model dynamic queueing systems leads to serious error even if deviation from stationarity is slight. The problem is compounded once we consider queueing networks. Understanding long-term and transient behaviour of such queues can only be achieved with approximation methods or simulation. We now detail the QDC algorithm, a computationally efficient method for simulating queueing systems.

@@

QDC can be considered as a multiserver extension to an algorithm presented by \citet{lindley_theory_1952}. For a single server queueing system, the departure time of the $i$th customer is: $d_i = \max{(a_i, d_{i-1})} + s_i$, since the customer either waits for a server or the server waits for a customer. The algorithm (not the paper) was, surprisingly, not extended to multiserver systems until \citet{krivulin_recursive_1994}. However with each new customer $i$ the algorithm must search a growing $i+1$ length vector. This algorithm, therefore, scales poorly, with computational complexity $O(n^2)$, where $n$ is number of customers. \citet{kin_generalized_2010} adapted the original algorithm of \citet{kiefer_theory_1955} to an $O(nK)$ algorithm for multiserver tandem queues with blocking, that is $G/G/K/C$ queueing systems where $C$ is the maximum capacity number of customers in the queueing systems.

QDC can also be viewed as a computationally efficient solution to the set of equations presented in \citet[pg. 259]{sutton_bayesian_2011} for FCFS queueing systems. There is a single queue served by a fixed number of $K$ servers. The $i$th customer observes a set of times $b_i = \{ b_{ik} | k \in 1:K \}$ which represents the times when each server will next be available. The customer $i$ selects the earliest available server $p_i = \text{argmin}(b_i)$ from $b_i$. The departure time for the $i$th customer is, therefore, $d_i = \max({a_i, b_{p_i}}) + s_i$, since the server must wait for the customer or the customer must wait for the server. The QDC algorithm for a fixed number of servers (Algorithm \ref{alg:fixed_servers}) pre-sorts the arrival times. Rather than assigning a $b_i$ for each customer $i$ to form the matrix $\mathbf{b} \in \mathbf{M}^{n \times K}$, QDC considers $\mathbf{b}$ as a continually updated $K$ length vector representing the state of the system.

\begin{algorithm}[ht!]
\caption{QDC for fixed $K$}
\label{alg:fixed_servers}
\begin{algorithmic}[1]
\Function{QDC\_numeric}{$\mathbf{a} \in \mathbb{R}_{+}^{n}, \mathbf{s} \in \mathbb{R}_{+}^{n}, K \in \mathbb{N}$}
\State Sort $(\mathbf{a}, \mathbf{s})$ in terms of $\mathbf{a}$ (ascending)
\State Create vector $\mathbf{p} \in \mathbb{N}^{n}$.
\State Create vector $\mathbf{b} \in \mathbb{R}_{+}^{K}$. 
\State Create vector $\mathbf{d} \in \mathbb{R}_{+}^{n}$.
\State $b_k \leftarrow 0 \quad \quad \forall k \in 1:K $
\For{$i \in 1:n$}
  \State $p_i \leftarrow \text{arg min}(\mathbf{b}) $ \label{alg_line:search}
  \State $b_{p_i} \leftarrow \max(a_i, b_{p_i}) + s_i $
  \State $d_i \leftarrow b_{p_i}$ 
\EndFor 
\State Put ($\mathbf{a}, \mathbf{d}, \mathbf{p}$) back to original (input) ordering of $\mathbf{a}$ 
\State \Return ($\mathbf{d}, \mathbf{p}$)
\EndFunction
\end{algorithmic}
\end{algorithm}

This algorithm is simple and computationally efficient. At each iteration of the loop, we need only search $\mathbf{b}$, a $K$ length vector for the minimum element in code line \ref{alg_line:search}. In the language of DES, we consider $\mathbf{b}$ as the system state and $\mathbf{a}$ as the event list, which are all determined events. This differs from conventional DES approaches to modelling queueing systems where the queue length is the system state, and both $\mathbf{a}$ and $\mathbf{d}$ constitute the event list, where the events of $\mathbf{a}$ are determined and the events of $\mathbf{d}$ are continually updated and therefore contingent.

Algorithm \ref{alg:fixed_servers} can simulate any queue of the form $G(t)/G(t)/K/\infty/n/FCFS$ where $K$ and $n$ can be made arbitrarily large. Furthermore, the inter-arrival and service distributions can be of completely general form and even have a dependency structure between them. Since the arrival and service times are supplied by the user rather than sampled in-situ, the algorithm ``decouples" statistical sampling from queue computation. This frees the user to simulate queues of arbitrarily complex $f_{\delta, \mathbf{s}}$, where $K$ is fixed.

\subsection{Changing number of servers}

@@

We represent the number of open servers throughout the day as a step function. Time is on the positive real number line and is partitioned by $L$ knot locations $\mathbf{x} = (x_1, \cdots, x_L ) \in \mathbb{R}^L_{+}$ into $L+1$ epochs $(0, x_1], (x_1, x_2], \cdots, (x_L, \infty)$. The number of open servers in each epoch is represented by a $L+1$ length vector $\mathbf{y} = (y_1, \cdots, y_{L+1} ) \in \mathbb{N}^{L+1}_0$. If we assume that none of the service times $\mathbf{s}$ span the length of more than one epoch $(x_{l}, x_{l+1}]$, formally

\begin{align}
\forall i \quad \left[ s_i < \min({ x_{l+1} - x_{l} | l \in 1:L }) \right] \label{eq:condition},
\end{align}

then we need to consider a change in state over at most 1 knot location. This step function is determined input by the user. Like the arrival and service times $\mathbf{(a,s)}$ it is changeable by the user before the simulation but not during the simulation.

@@

Clearly, QDC and its implementation \pkg{queuecomputer} are a more computationally efficient way to simulate queueing systems of the form $G(t)/G(t)/K/\infty/M/FCFS$ than conventional DES algorithms implemented by \pkg{simpy} and \pkg{simmer}.

>>>>

\documentclass[article]{jss}
\usepackage{thumbpdf,lmodern} 
\graphicspath{{Figures/}}

\usepackage{amsmath,amssymb,array}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{algorithm}
\usepackage{rotating}

\author{Anthony  Ebert\\Queensland University of Technology\\ACEMS \And 
        Paul Wu\\Queensland University of Technology\\ACEMS \AND
        Kerrie Mengersen\\Queensland University of Technology\\ACEMS \And
        Fabrizio Ruggeri\\CNR-IMATI\\Queensland University of Technology\\ACEMS}
\title{Computationally Efficient Simulation of Queues: The \proglang{R} Package \pkg{queuecomputer}}

\Shorttitle{\pkg{queuecomputer}: Computationally Efficient Simulation of Queues}

\Plainauthor{Anthony Ebert, Paul Wu, Kerrie Mengersen, Fabrizio Ruggeri}
\Plaintitle{Computationally Efficient Simulation of Queues: The R Package queuecomputer}

\Abstract{ Large networks of queueing systems model important
  real-world systems such as MapReduce clusters, web-servers,
  hospitals, call centers and airport passenger terminals. To model
  such systems accurately, we must infer queueing parameters from
  data. Unfortunately, for many queueing networks there is no clear
  way to proceed with parameter inference from data. Approximate
  Bayesian computation could offer a straightforward way to infer
  parameters for such networks if we could simulate data quickly
  enough.

  We present a computationally efficient method for simulating from a
  very general set of queueing networks with the \proglang{R} package
  \pkg{queuecomputer}. Remarkable speedups of more than 2 orders of
  magnitude are observed relative to the popular DES packages
  \pkg{simmer} and \pkg{simpy}. We replicate output from these
  packages to validate the package.

  The package is modular and integrates well with the popular
  \proglang{R} package \pkg{dplyr}. Complex queueing networks with
  tandem, parallel and fork/join topologies can easily be built with
  these two packages together. We show how to use this package with
  two examples: a call center and an airport terminal.  }

\Keywords{queues, queueing theory, discrete event simulation, 
  operations research, approximate Bayesian computation, \proglang{R}}

\Plainkeywords{queues, queueing theory, discrete event simulation, 
  operations research, approximate Bayesian computation, R}

%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

\Address{
  Anthony Ebert\\
  School of Mathematical Sciences\\
  Science and Engineering Faculty\\
  Queensland University of Technology\\
  Brisbane Queensland 4000, Australia\\
  E-mail: \email{ac.ebert@qut.edu.au}\\
  URL: \url{https://acems.org.au/our-people/anthony-ebert}
}

@@

Performance measures for a given queueing system can often only be
derived through simulation. Queues are usually simulated with discrete
event simulation \citep[DES;][p.~226]{insua2012bayesian}. In DES
changes in state are discontinuous. The state is acted upon by a
countable list of events at certain times which cause the
discontinuities. If the occurrence of an event is independent of
everything except simulation time it is determined; otherwise, it is
contingent \citep{nance1981time}.

Popular DES software packages are available in many programming
languages including: the \proglang{R} \citep{Rproglang} package
\pkg{simmer} \citep{Rpkg_simmer}, the \proglang{Python}
\citep{van2011python} package \pkg{simpy} \citep{Ppkg_simpy} and the
\proglang{Java} \citep{gosling2000java} package \pkg{JMT}
\citep{Jpkg_JMT}. DES packages are often so expressive that they can
be considered languages in their own right, indeed the programming
language \proglang{SIMULA} \citep{dahl1966simula} is a literal example
of this.

The \proglang{R} package \pkg{queuecomputer}
\citep{Rpkg_queuecomputer}, which is available from the Comprehensive
\proglang{R} Archive Network (CRAN) at
\url{https://CRAN.R-project.org/package=queuecomputer}, implements an
algorithm that can easily be applied to a wide range of queueing
systems and networks of queueing systems. It is vastly more
computationally efficient than existing approaches to DES. We term
this new computationally efficient algorithm queue departure
computation (QDC). Computational efficiency is important because if we
can simulate from queues quickly, then we can embed a queue simulation
within an approximate Bayesian computation (ABC) algorithm
\citep{sunnaker_approximate_2013} and estimate queue parameters for
very complicated queueing models in a straightforward manner.

In Section~\ref{sec:queueing} we review the literature on queueing
theory and develop notation used throughout this paper. In
Section~\ref{sec:QDC} we present the QDC algorithm and compare it to
DES. We demonstrate usage of the package in
Section~\ref{sec:usage}. Details of implementation and usage are
discussed in Section~\ref{sec:Implementation}. The package is
validated in Section~\ref{sec:Validation} by replicating results from
DES packages \pkg{simpy} and \pkg{simmer}. We compare computed
performance measures from the output of a \pkg{queuecomputer}
simulation to theoretical results for $M/M/2$ queueing systems. We
benchmark the package in Section~\ref{sec:Benchmark} and compare
computation time with \pkg{simpy} and \pkg{simmer}. Examples in
Section~\ref{sec:Examples} are used to demonstrate how the package can
be used to simulate a call center and an international airport
terminal.

@@

Queueing theory is the study of queueing systems and originated from
the work of Agner Krarup Erlang in 1909 to plan infrastructure
requirements for the Danish telephone system
\citep[p.~2]{thomopoulos2012fundamentals}. An overview of the notation
and definitions used in this paper is given in
Table~\ref{tab:notation}.

A queueing system is defined as follows. Each customer
$i = 1,2,\ldots$ has an arrival time $a_i$ (or equivalently an
inter-arrival time $\delta_i = a_i - a_{i-1}$, $a_0 = 0$) and an
amount of time they require with a server, called the service time
$s_i$. Typically a server can serve only one customer at a time. A
server which is currently serving another customer is said to be
unavailable, a server without a customer is available. If all servers
are unavailable when a customer arrives then customers must wait in
the queue until a server is available. Detailed introductions to
queueing systems can be found in standard texts such as
\citet{bhat2015introduction}.

The characteristics of a queueing system are expressed with the notation of \citet{kendall1953stochastic}. This notation has since been extended to six characteristics:
\begin{itemize}
\item $f_{\mathbf{\delta}}$, inter-arrival distribution;
\item $f_{\mathbf{s}}$, service distribution; 
\item $K$, number of servers $\in \mathbb{N}$;
\item $C$, capacity of system $\in \mathbb{N}$; 
\item $n$, customer population $\in \mathbb{N}$; and
\item $R$, service discipline.
\end{itemize}

Choices for inter-arrival and service distributions are denoted by
``M'' for exponential and independently distributed, ``GI'' for general
and independently distributed and ``G'' for general without the
independence assumption. The capacity of the system $C$ refers to the
maximum number of customers within the system at any one
time\footnote{If the system is at full capacity and new customers
  arrive, new customers leave the system immediately without being
  served.}. Customers are within the system if they are being served
or waiting in the queue. The customer population $n$ is the total
number of customers including those outside of the system (yet to
arrive or already departed). The service discipline $R$ defines how
customers in the queue are allocated to available servers. The most
common service discipline is first come first serve (FCFS). To specify
a queueing system, these characteristics are placed in the order given
above and separated by a forward slash ``/''.

The simplest queueing system is exponential in distribution for both
the inter-arrival
$\delta_i \overset{\text{iid}}{\sim} \mathrm{exp}(\lambda) ~ \forall i \in
1:n$ and service processes
$s_i \overset{\text{iid}}{\sim} \mathrm{exp}(\mu) ~ \forall i \in 1:n$, where
$\lambda$ and $\mu$ are exponential rate parameters. Additionally, $K$
is set to 1, $C$ and $n$ are infinite, and $R$ is FCFS. It is denoted
by $M/M/1/ \infty / \infty / \mathit{FCFS}$, which is shortened to $M/M/1$.

Parameter inference for this system was considered first by
\citet{clarke1957maximum}, estimators were derived from the likelihood
function. This likelihood is later used by
\citet{muddapur1972bayesian} to derive the joint posterior
distribution. Bayesian inference for queueing systems is summarized in
detail by \citet{insua2012bayesian}.

\begin{sidewaystable}
\centering
% \resizebox{0.9\textwidth}{!}{
\begin{tabular}{llp{14cm}}
\hline
Type & Notation & Definition \\
\hline

Queue specification

& $\lambda$ & Rate parameter of exponential inter-arrival distribution $f_\delta = \text{exp}(\lambda)$ for $M/M/K$ queue. \\
& $\mu$ & Rate parameter of exponential service distribution $f_{\mathbf{s}} = \text{exp}(\mu)$ for $M/M/K$ queue.  \\
& $\rho := \frac{\lambda}{K \mu}$ & Traffic intensity, defined only for $M/M/K$ queues. \\ 
& $\theta_{I}$ & Parameters of arrival and service joint distribution for QDC, $(\mathbf{a}, \mathbf{s}) \sim f_{\mathbf{a}, \mathbf{s}}(\cdot | \theta_I )$. \\

& $K$ & Number of servers. \\
& $C$ & Capacity of system.  \\
& $n$ & Total number of customers. \\
& $R$ & Service discipline of queue. \\
& FCFS & First come first serve (type of service discipline). \\
& $L$ & Number of knot locations for server changes. \\

Input/output
& $\mathbf{a} = (a_1, \ldots, a_n )$ & Arrival process, where $a_i$ is the time at which the $i$th customer arrives at the queue. \\

& $\mathbf{\delta} = (\delta_1, \ldots, \delta_n )$ & Inter-arrival process, where $\delta_i$ is $a_i - a_{i-1} \quad \forall i \in 1:n$ and $\delta_1 = a_1$. \\

& $\mathbf{s} = (s_1, \ldots, s_n )$ & Service process, where $s_i$ is the service time of the $i$th customer. \\

& $\mathbf{d} = (d_1, \ldots, d_n )$ & Departure process, where $d_i$ is the time at which the $i$th customer leaves the queue after being served. \\

& $\mathbf{p} = (p_1, \ldots, p_n )$ & Server process, where $p_i$ is the server who served the $i$th customer. \\

& $\mathbf{b} = (b_1, \ldots, b_K )$ & This vector represents the time at which each server $1:K$ will next be free. We consider this vector to be the state of the system. \\

& $\mathbf{x} = (x_1, \ldots, x_L )$ & Change times for number of open servers. \\

& $\mathbf{y} = (y_1, \ldots, y_{L+1} )$ & Number of open servers in each epoch. \\

Performance measures

& $N(t)$ & Number of customers in the system at time $t$. \\
& $\bar{B}$ & Average number of busy servers, referred to as ``resource utilization''. \\
& $\bar{w}$ & Average waiting time per customer. \\

\hline
\end{tabular}
% }
\caption{Notation and definitions.}\label{tab:notation}
\end{sidewaystable}

@@

In the case of a $M/M/K$ system equilibrium distributions for
performance measures are derived analytically, they are found in
standard queueing theory textbooks \citep{lipsky2008queueing,
  thomopoulos2012fundamentals}. For instance, the limit probability of
$N$ customers in the system $\Prob(N)$ is
%
\begin{align}
\Prob(0) &= \left[\frac{ (K \rho)^K}{K! (1 - \rho)} + 1 + \sum_{i=1}^{K-1} \frac{(K \rho)^i}{i!}   \right]^{-1}, \nonumber \\
\Prob(N) &= \begin{cases} 
\Prob(0) \frac{(K \rho)^n }{N!} & \quad  N \leq K,\\
\Prob(0) \frac{(K \rho)^n }{K! K^{N-K}} & \text{otherwise} ,
\end{cases} \label{eq:pn}
\end{align}
%
where $\rho$, the resource utilization, is defined as
$\frac{\lambda}{K \mu}$. For an $M/M/K$ system this is equal to the
expected number of busy servers divided by the total number of servers
$\frac{\E(B)}{K}$ \citep[p.~451]{cassandras2009introduction}. The
expected number of customers in the system is
\citep{bhat2015introduction}
%
\begin{align}
\E(N) = K\rho + \frac{\rho (K\rho)^K P(0)}{K!(1-\rho)^2} \label{eq:en},
\end{align}
%
and the expected waiting time is
%
\begin{align}
\E(\mathbf{w}) = \frac{(K \rho)^K P(0)}{K! K \mu(1-\rho)^2} \label{eq:ew}.
\end{align}
%
If the parameters of $f_\delta$ and $f_{\mathbf{s}}$ are uncertain,
then we must turn to predictive distributions for estimates of
performance measures, which are computed analytically for $M/M/K$
queues (Equations~\ref{eq:en} and~\ref{eq:ew}). Predictive
distributions of performance measures using Bayesian posterior
distributions are derived by \citet{armero1994bayesian,
  armero1999dealing}.

@@

In a tandem queueing network, customers traverse an ordered series of
queues before departing the system. Real examples of such systems
include airport terminals, internet services and manufacturing
systems. In a parallel network, customers are partitioned into
different $\mathbf{(a,s)}$ to be seen by separate queueing systems. In
a fork/join network a task (another term for customer) is forked into
a number of subtasks which are to be completed by distinct parallel
servers. The difference from the parallel network is that the task can
only depart the system once all subtasks have arrived at the join
point.

@@

In general, analytic solutions do not exist for dynamic queueing
systems \citep{malone1995dynamic,
  worthington2009reflections}. \citet{green1991some} showed that using
stationary queueing systems to model dynamic queueing systems leads to
serious error even if deviation from stationarity is slight. The
problem is compounded once we consider queueing
networks. Understanding long-term and transient behavior of such
queues can only be achieved with approximation methods or
simulation. We now detail the QDC algorithm, a computationally
efficient method for simulating queueing systems.

@@

QDC can be considered as a multiserver extension to an algorithm
presented by \citet{lindley_theory_1952}. For a single server queueing
system, the departure time of the $i$th customer is:
$d_i = \max{(a_i, d_{i-1})} + s_i$, since the customer either waits
for a server or the server waits for a customer. The algorithm (not
the paper) was, surprisingly, not extended to multiserver systems
until \citet{krivulin_recursive_1994}. However with each new customer
$i$ the algorithm must search a growing $i+1$ length vector. This
algorithm, therefore, scales poorly, with computational complexity
$O(n^2)$, where $n$ is the number of
customers. \citet{kin_generalized_2010} adapted the original algorithm
of \citet{kiefer_theory_1955} to an $O(nK)$ algorithm for multiserver
tandem queues with blocking, that is $G/G/K/C$ queueing systems where
$C$ is the maximum capacity number of customers in the queueing
systems.

QDC can also be viewed as a computationally efficient solution to the
set of equations presented in \citet[p.~259]{sutton_bayesian_2011} for
FCFS queueing systems. There is a single queue served by a fixed
number of $K$ servers. The $i$th customer observes a set of times
$b_i = \{ b_{ik} | k \in 1:K \}$ which represents the times when each
server will next be available. The customer $i$ selects the earliest
available server $p_i = \text{argmin}(b_i)$ from $b_i$. The departure
time for the $i$th customer is, therefore,
$d_i = \max({a_i, b_{p_i}}) + s_i$, since the server must wait for the
customer or the customer must wait for the server. The QDC algorithm
for a fixed number of servers (Algorithm~\ref{alg:fixed_servers})
pre-sorts the arrival times. Rather than assigning a $b_i$ for each
customer $i$ to form the matrix
$\mathbf{b} \in \mathbf{M}^{n \times K}$, QDC considers $\mathbf{b}$
as a continually updated $K$ length vector representing the state of
the system.

\begin{algorithm}[t!]
\caption{QDC for fixed $K$.}
\label{alg:fixed_servers}
\begin{algorithmic}[1]
\Function{QDC\_numeric}{$\mathbf{a} \in \mathbb{R}_{+}^{n}, \mathbf{s} \in \mathbb{R}_{+}^{n}, K \in \mathbb{N}$}
\State Sort $(\mathbf{a}, \mathbf{s})$ in terms of $\mathbf{a}$ (ascending).
\State Create vector $\mathbf{p} \in \mathbb{N}^{n}$.
\State Create vector $\mathbf{b} \in \mathbb{R}_{+}^{K}$. 
\State Create vector $\mathbf{d} \in \mathbb{R}_{+}^{n}$.
\State $b_k \leftarrow 0 \quad \quad \forall k \in 1:K $
\For{$i \in 1:n$}
  \State $p_i \leftarrow \text{arg min}(\mathbf{b}) $ \label{alg_line:search}
  \State $b_{p_i} \leftarrow \max(a_i, b_{p_i}) + s_i $
  \State $d_i \leftarrow b_{p_i}$ 
\EndFor 
\State Put ($\mathbf{a}, \mathbf{d}, \mathbf{p}$) back to original (input) ordering of $\mathbf{a}$.
\State \Return ($\mathbf{d}, \mathbf{p}$)
\EndFunction
\end{algorithmic}
\end{algorithm}

This algorithm is simple and computationally efficient. At each iteration of the loop, we need only search $\mathbf{b}$, a $K$ length vector for the minimum element in code line~\ref{alg_line:search}. In the language of DES, we consider $\mathbf{b}$ as the system state and $\mathbf{a}$ as the event list, which are all determined events. This differs from conventional DES approaches to modeling queueing systems where the queue length is the system state, and both $\mathbf{a}$ and $\mathbf{d}$ constitute the event list, where the events of $\mathbf{a}$ are determined and the events of $\mathbf{d}$ are continually updated and therefore contingent.

Algorithm~\ref{alg:fixed_servers} can simulate any queue of the form $G(t)/G(t)/K/\infty/n/\mathit{FCFS}$ where $K$ and $n$ can be made arbitrarily large. Furthermore, the inter-arrival and service distributions can be of completely general form and even have a dependency structure between them. Since the arrival and service times are supplied by the user rather than sampled in-situ, the algorithm ``decouples'' statistical sampling from queue computation. This frees the user to simulate queues of arbitrarily complex $f_{\delta, \mathbf{s}}$, where $K$ is fixed.

\subsection{Changing the number of servers}

@@

We represent the number of open servers throughout the day as a step
function. Time is on the positive real number line and is partitioned
by $L$ knot locations
$\mathbf{x} = (x_1, \ldots, x_L ) \in \mathbb{R}^L_{+}$ into $L+1$
epochs $(0, x_1], (x_1, x_2], \ldots, (x_L, \infty)$. The number of
open servers in each epoch is represented by a $L+1$ length vector
$\mathbf{y} = (y_1, \ldots, y_{L+1} ) \in \mathbb{N}^{L+1}_0$. If we
assume that none of the service times $\mathbf{s}$ span the length of
more than one epoch $(x_{l}, x_{l+1}]$, formally
%
\begin{align}
\forall i \quad \left[ s_i < \min({ x_{l+1} - x_{l} | l \in 1:L }) \right] \label{eq:condition},
\end{align}
%
then we need to consider a change in state over at most 1 knot
location. This step function is determined input by the user. Like the
arrival and service times $\mathbf{(a,s)}$ it is changeable by the
user before the simulation but not during the simulation.

@@

Clearly, QDC and its implementation \pkg{queuecomputer} are a more computationally efficient way to simulate queueing systems of the form $G(t)/G(t)/K/\infty/M/\mathit{FCFS}$ than conventional DES algorithms implemented by \pkg{simpy} and \pkg{simmer}.
